Research Notes

### 2018-09-09 writeups/notes.tex
check in writeup progress

\section{Linear Amplification}

One can solve \autoref{eq:shear_lin_uz}

\section{Breaking Criterion}

If the fluid is not strongly stratified $\ptd{u_0}{z}, \pd{u_0}{z}\frac{1}{H}
\ll \frac{k_xN^2}{\omega_{cr}}$ then the singularity is largely induced by
$\omega_{cr} \to 0$ (note that the condition becomes easier to satisfy as we
approach the singularity). $\omega_{cr} = 0$, is exactly the \emph{corotation
resonance}, and one expects that as (note that if $k_z \gg k_x$ then the
dispersion relation is $\omega_{cr} \approx Nk_x/k_z$):
\begin{equation}
    k_z(z) \approx \frac{Nk_x}{\omega_{cr}(z)}
        \sim \frac{\omega_{cr}(z)}{\abs*{u_{1z}}},
        \label{eq:corot_nonlin_cond}
\end{equation}
that nonlinear effects should kick in.

One advantage of a shear flow is that, instead of requiring $2H$ of vertical
height for amplitude growth, we can just choose $u_0 =
\frac{\omega}{k_x}\frac{z}{z_{\max}}$ to meet the corotation citerion at $z =
z_{\max}$. Consider $\pd{u_0}{z} \sim \frac{\omega}{k_xz_{\max}} \approx
\frac{N}{k_zz_{\max}}$. This both satisfies $\pd{u_0}{z} \frac{1}{H} =
\frac{N}{k_zz_{\max}H} \ll \frac{k_xN^2}{\omega_{cr}}$ (as $k_z^2z_{\max}H \ll
1$), and $\ptd{u_0}{z} = 0 \ll \frac{k_xN^2}{\omega_{cr}}$. We can even induce
the corotation resonance well below $z_{\max}$ and still be perfectly within the
regime of validity of our assumption.

A second advantage of a shear flow is that we can force at very small amplitudes
and still observe resonances, since the resonance is driven by the convergence
of $\omega_{cr} \to 0$ (purely coordinate driven) instead of $u_{1z} \sim
v_{ph}$ (dependent on wave amplitude).

The width of the dissipation zone can be estimated by noticing $u_0k_x = \omega
\pm \sqrt{k_xNu_{1z}}$ from the nonlinearity criterion
\autoref{eq:corot_nonlin_cond}. Thus, the width of $u_0$ that incurs nonlinear
effects is roughly of order $2\sqrt{Nu_{1z} / k_x}$, which has the right scaling
behavior (a stronger perturbation should be resonant farther away).

>>>>

\section{Reflection/Transmission in Linear Theory}

One can solve \autoref{eq:shear_lin_uz} via \emph{Frobenius Theory}, but if we
consider only $\omega_{cr} \to 0$, we obtain a very simple power series
solution. In $\omega_{cr} \to 0$, we have only leading order contributions
\begin{equation}
    \ptd{u_{1z}}{z} + \frac{k_x^2N^2}{\omega_{cr}^2}u_{1z} = 0.
\end{equation}
Consider a power series solution $u_{1z} \propto \omega_{cr}^\alpha$, where
$\omega_{cr}(z) = \omega - k_xu_0(z)$. Note that very near the critical level,
$\pd{u_0}{z}\delta z k_x$ where $\delta z = z - z_c, \omega_{cr}(z_c) = 0$. We
thus have
\begin{align}
    \p*{\pd{\omega_{cr}}{z}}^2 \alpha\p*{\alpha - 1} + k_x^2N^2 &= 0,\\
    \alpha &= \frac{1}{2}\pm i\sqrt{\frac{N^2}{\p*{\pd{u_0}{z}}^2} -
        \frac{1}{4}}.
\end{align}
The classic Booker \& Bretherton paper notates this result $u_{1z} \propto
\p*{\omega_{cr}(z)}^{\frac{1}{2} \pm i\mu} \propto (\delta z)^{\frac{1}{2} \pm
i\mu}$.

Now, consistent with a perturbation that vanishes as $t \to -\infty$
(alternatively, this is just the Landau prescription for deforming below a
singularity on the real line), we must interpret $\omega_{cr}$ as having a
slight positive imaginary component. Given this, the expression for $u_{1z}(z)$
is multivalued, and we must specify a branch cut. For $\Arg\p*{z - z_c} = 0$, or
intuitively when we are above the critical layer, choose branch such that
\begin{equation}
    \p*{z - z_c}^{\frac{1}{2} \pm i\mu} \equiv
        \abs*{z - z_c}^{1/2\pm i\mu}.
\end{equation}
It is then clear that for $\Arg\p*{z - z_c} = -\pi$, intuitively corresponding
to being below the critical layer (note the arg is $-\pi$ since $z_c$ is
infinitesimally above the real axis), our branch cut enforces
\begin{equation}
    \p*{z - z_c}^{\frac{1}{2} \pm i\mu} =
        \abs*{z - z_c}^{\frac{1}{2} \pm i\mu}e^{-\frac{i\pi}{2}}e^{\mp \mu \pi}.
\end{equation}
Lastly, we interpret $\abs*{z - z_c}^{\pm i\mu} = e^{\pm i\mu \ln \abs*{z -
z_c}}$ to have the form of a positive/negative $k_z$ value. Interestingly, this
means that $e^{-i\mu \ln \abs*{z - z_c}}$ corresponds to the \emph{positive
group velocity}, which is $e^{\mu \pi}$ magnified compared to the $\Arg\p*{z -
z_c} = 0$ case, and $e^{2\mu \pi}$ magnified over the negative group velocity
component below $z_c$.

The interpretation we choose for this is that an incoming wavepacket moving with
positive group velocity inbound on $z_c$ is transmitted with amplitude $e^{-\mu
\pi}$ and reflected with amplitude $e^{-2\mu \pi}$. The remainder is interpreted
as being \emph{absorbed} by the critical layer.

\section{Validity of the Linear Approximation}

The real physics is expected to deviate from our computation near the critical
layer owing primarily to two effects: nonlinear advective terms and diffusivity.
Hazel 1967 found numerically that including diffusive tersm in the equations of
motion does not change the attenuation factor; even in the presence of
dissipation no reflection/transimission are observed. This makes sense;
examining the equations of motion, the effect of an extra $\pd{}{t} \to \pd{}{t}
- \nu \nabla^2$ is to take $\omega_{cr} \to \omega_{cr} + i\nu k_x^2 - i\nu
\ptd{}{z}$, giving the singularity at $\omega_{cr} + i\nu k_x^2 = 0$ the
positive imaginary component we before asserted.

On the other hand, other studies e.g.\ Brown \& Stewartson 1980, Winters \&
d'Asaro 1994, find that once the wave amplitude becomes nonnegligible, sizable
reflection can be found. In contrast, Ogilvie \& Barker 2010 find that in
geometrically focused breaking, no reflection is observed, despite the nonlinear
effects being stronger than the dissipitive ones. The study of how nonlinear and
diffusive effects compete is then nontrivial.
